Equalizer Programming Guide

### 2009-05-30 Developer/ProgrammingGuide/paper.tex
PG: Update documentation related Equalizers.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.5.7 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
    1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
    1.5.5   & Mar 10, 2009 & Updated to reflect new AGL event dispatch\\
    1.5.6   & Mar 23, 2009 & Documented Layout API\\
    1.5.7   & Apr 17, 2009 & Updated Layout API (multiple observers)\\
    \hline
  \end{tabularx}}
\clearpage

@@

\subsection{\label{sLoadBalancing}Load-Balancing}

While pixel and stereo compounds are naturally load-balanced, 2D, DB and
DPlex compounds might need load-balancing to provide optimal
performance. Using a load-balancer is transparent to the application,
and can be used on any application for 2D and DPlex, and on most
applications for DB load-balancing. Some applications supporting DB
decomposition do not support dynamic updates of the database range, and
therefore cannot be used with DB load-balancing.

Using a 2D or DB load-balancer will adapt the 2D split or database range
automatically each frame. The 2D load-balancer exists in three flavors:
\textsf{2D} using tiles, \textsf{VERTICAL} using columns and
\textsf{HORIZONTAL} using rows. The current load-balancing algorithm for
2D and DB is a technology preview and will improve in future versions.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/lb.pdf}
  {\caption{\label{fLoadBalancing}\small 2D load-balancing}}
\end{wrapfigure}
2D load-balancing increases the framerate over a static decomposition in
virtually all cases. It works best if the application data is relatively
uniformly distributed in screen-space.

DPlex compounds require a smoothing of the framerate at the destination
channel, otherwise the framerate will become periodically faster and
slower. Using a DPlex load-balancer will smoothen the swap buffer rate
on the destination window for optimal user experience.

@@

The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\label{fModel}Simplified execution model}}
\end{wrapfigure}
All rendering threads in the configuration run asynchronously to the
application's main thread. Depending on the configuration's latency,
they can fall $n$ frames behind the last frame finished by the
application thread. A latency of one frame is usually not perceived by
the user, but can increase rendering performance substantially.

@@

A layout is the grouping of logical views. It is used by one or more
canvases. For all given layouts, Equalizer creates destination channels
when the configuration file is loaded. These destination channels can be
referenced by compounds to configure scalable rendering.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/layout.png}
  {\caption{\label{fLayout}A layout with four views}}
\end{wrapfigure}
A view is a logical view of the application data, in the sense used by
the Model-View-Controller pattern. It can be a scene, viewing mode,
viewing position, or any other representation of the application's data.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
  {\caption{A display wall using a six-segment canvas with a two-view layout.}}
\end{figure}

@@

\subsubsection{Synchronization - Swapbarrier}
Compounds may have a swapbarrier to synchronize the buffer swap of
multiple channels. Windows with a swapbarrier of the same name enter a
barrier before executing the swap buffer task. Before entering the
barrier, \textsf{Window::finish} is called to ensure that all OpenGL
commands have been executed.

Note that this software swap barrier is not accurate enough for
edge-blended projection system. Such installations typically use a
hardware synchronization, e.g., nVidia G-Sync cards. It is however
sufficient for display walls made out of LCD's and for some
non-immersive projection systems.

@@

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/uml}
  {\caption{\label{fUml}UML diagram eqPly and relevant Equalizer classes.}}
\end{figure}

>>>>

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.5.8 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
    1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
    1.5.5   & Mar 10, 2009 & Updated to reflect new AGL event dispatch\\
    1.5.6   & Mar 23, 2009 & Documented Layout API\\
    1.5.7   & Apr 17, 2009 & Updated Layout API (multiple observers)\\
    1.5.8   & May 30, 2009 & Added \sref{sEqualizers}, updated \sref{sLoadBalancing} and \sref{sSwapBarrier}\\
    \hline
  \end{tabularx}}
\clearpage

@@

\subsection{\label{sLoadBalancing}Automatic Runtime Adjustments}

Some scalable rendering parameters have to be updated at run-time to
provide optimal performance or influence other scalable rendering
features. These adjustments are often referred to as load-balancing, but
are called \textsf{equalizers}, since their functionality is not limited
to load-balancing in Equalizer.

The server provides a number of equalizers, which automatically update
certain parameters of compounds based on run-time information. They
balance the load of 2D and DB compounds, optimally distribute render
resources for multi-view configurations, adapt the resolution to provide
a constant framerate or zoom images to allow monitoring of another view.

Equalizer are described in more detail in \sref{sEqualizers}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\label{fModel}Simplified execution model}}
\end{wrapfigure}
The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

All rendering threads in the configuration run asynchronously to the
application's main thread. Depending on the configuration's latency,
they can fall $n$ frames behind the last frame finished by the
application thread. A latency of one frame is usually not perceived by
the user, but can increase rendering performance substantially.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/layout.png}
  {\caption{\label{fLayout}A layout with four views}}
  \vspace{-4ex}
\end{wrapfigure}
A layout is the grouping of logical views. It is used by one or more
canvases. For all given layouts, Equalizer creates destination channels
when the configuration file is loaded. These destination channels can be
referenced by compounds to configure scalable rendering.

@@

A view is a logical view of the application data, in the sense used by
the Model-View-Controller pattern. It can be a scene, viewing mode,
viewing position, or any other representation of the application's data.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
  {\caption{A display wall using a six-segment canvas with a two-view layout}}
\end{figure}

@@

\subsubsection{\label{sSwapBarrier}Synchronization - Swapbarriers}
Compounds may have a software or hardware swapbarrier to synchronize the
buffer swap of multiple channels. A software swapbarrier is configured
by giving it a name. Windows with a swapbarrier of the same name
synchronize with each other before executing the swap buffer
task. Before entering the barrier, \textsf{Window::finish} is called to
ensure that all OpenGL commands have been executed.

Note that a software swap barrier is not accurate enough for
edge-blended projection system. Such installations typically use a
hardware synchronization, e.g., nVidia G-Sync cards. It is however
sufficient for display walls made out of LCD's and for some
non-immersive projection systems.

@@

A hardware swapbarrier is configured by setting the \textsf{NV\_group}
and \textsf{NV\_barrier} parameters. These parameters follow the
\textsf{NV\_swap\_group} extension, which synchronizes all windows bound
to the same group on a single machine, and all groups bound to the same
barrier across systems.

\subsubsection{\label{sEqualizers}Adaptation - Equalizers}

Equalizers are used to update compound parameters based on run-time
data. They are attached to a compound (sub-)tree, on which they
operate. The Equalizer distribution contains numerous example
configuration files using equalizers.

\paragraph{Load Equalizer}
\vspace{-2ex}\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/lb.pdf}
  {\caption{\label{fLoadBalancing}\small 2D load-balancing}}
\end{wrapfigure}
While pixel and stereo compounds are naturally load-balanced, 2D and DB
compounds often need load-balancing for optimal rendering
performance. Using a load equalizer is transparent to the application,
and can be used on any application for 2D, and on most applications for
DB load-balancing. Some applications do not support dynamic updates of
the database range, and therefore cannot be used with DB load-balancing.

Using a 2D or DB load-balancer will adapt the 2D split or database range
automatically each frame. The 2D load-balancer exists in three flavors:
\textsf{2D} using tiles, \textsf{VERTICAL} using columns and
\textsf{HORIZONTAL} using rows.

2D load-balancing increases the framerate over a static decomposition in
virtually all cases. It works best if the application data is relatively
uniformly distributed in screen-space. A damping parameter can be used
to fine-tune the algorithm.

\paragraph{View Equalizer}
Depending on the model position and data structure, each segment of a
multi-display system has a different rendering load. The segment with
the biggest load determines the overall performance when using a static
assignment of resources to segments. The view equalizer analyzes the
load of all segments, and adapts the resource usage each frame. It
equalizes the load on all segments of a view.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-segment
      load-balancing for two segments using eight GPU's}}
\end{wrapfigure}
\fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
  AG, 2008} illustrates this process. On the left side, a static
assigment of resources to display segments is used. The right-hand
segment has a higher load than the left-hand segment, causing
sub-optimal performance. The configuration on the left uses a view
equalizer, which assigns two GPU's to the left segment and four GPU's to
the right segment, which leads to optimal performance for this model and
camera position.

The view equalizer can also use resources from another display resource,
if this resource has little rendering load by itself. It is therefore
possible to improve the rendering performance of a multi-display system
without any additional resources. This is particulary useful for
installations with a higher number of displays where the rendering load
is typically in a few segments only, e.g., for a CAVE.

\begin{wrapfigure}{l}{.382\textwidth} 
    \includegraphics[width=.382\textwidth]{images/viewLB.png}
    {\caption{\label{fViewLoadBalancing2}\small Cross-segment
        load-balancing for a CAVE}}
\end{wrapfigure}
\fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
by five GPU's. The front and left segments show the model and have a
significant rendering load. The view equalizer assigns the GPU's from
the top, bottom and right wall for rendering the left and front wall in
this configuration.

Cross-segment load-balan\-cing is configured hierarchically. On the top
compound level, a view equalizer assigns resources to each of its
children, so that the optimal number of resources is used for each
segment. On the next level, a load equalizer on each child computes the
resource distribution within the segment, taking the resource usage
given by the view equalizer into account.

\paragraph{Framerate Equalizer}
Certain configurations, in particular DPlex compounds, require a
smoothing of the framerate at the destination channel, otherwise the
framerate will become periodically faster and slower. Using a framerate
equalizer will smoothen the swap buffer rate on the destination window
for optimal user experience.

\paragraph{DFR Equalizer}
Dynamic Frame Resolution (DFR) trades rendering performance for visual
quality. The rendering for a channel is done at a different resolution
than the native channel resolution to keep the framerate constant. The
DFR equalizer adapts the zoom of a channel, based on the target and
current framerate. It is typically used for fill-rate bound
applications, such as volume rendering and raytracing.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
  {\caption{\label{fDFR}\small Dynamic Frame Resolution}}
\end{wrapfigure}
\fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
  Science Institute, University of Freiburg, Germany} shows DFR for
volume rendering. To achieve 10 frames per second, the model is rendered
at a lower resolution, and upscaled to the native resolution for
display. The rendering quality is slightly degraded, while the rendering
performance remains interactive. When the application is idle, it can
render a full-resolution view.

The dynamic frame resolution is not limited to down-scaling the
rendering resolution, it will also up-scale the resolution if the
parameters allow for it. Upscaled rendering, which will downsample the
result for display, provides dynamic anti-aliasing at a constant
framerate.

\paragraph{Monitor Equalizer}
The monitor equalizer allows the observation of another view,
potentially made of multiple segments, in a different channel at a
different resolution. This is typically used to re-use the rendering of
a large-scale display on an operator station.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/monitorEq.pdf}
  {\caption{\label{fMonitorEq}\small Monitoring a projection wall}}
\end{wrapfigure}
The monitor equalizer adapts the frame zoom of the output frames used to
observe the rendering, depending on the destination channel size. The
output frames are downscaled on the GPU before readback, which results
in optimal performance.

\fig{fMonitorEq} shows a usage of the monitor equalizer. A two-segment
display wall is driven by a separate control station. The rendering
happens only on the display wall, and the control window receives the
correctly downscaled version of the rendering.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/uml}
  {\caption{\label{fUml}UML diagram eqPly and relevant Equalizer classes}}
\end{figure}
